\documentclass{wissdoc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{microtype}
\usepackage{paralist}
\usepackage{listings} 
\usepackage{rotating}
\usepackage[square, comma, sort&compress]{natbib}
%\lstset{numbers=left, numberstyle=\tiny, numbersep=5pt} 
\lstset{language=C}
%\lstset{xleftmargin=1cm}
\lstset{
%fancyvrb=true,
aboveskip=15pt,
captionpos=b,
%basicstyle=\small,
basicstyle=\ttfamily\footnotesize,
identifierstyle=\textbf,
numbers=left,
numberstyle=\ttfamily\tiny,
numbersep=10pt,
numberblanklines=false,
stepnumber=1,
tabsize=2,
postbreak=\space,
breaklines=true, % Zeilenumbruch
breakautoindent=true, % Bei Zeilenumbruch einrücken
frame=lines,
xleftmargin=0.05\textwidth,
linewidth=0.95\textwidth
}

%% Informationen für die PDF-Datei
\hypersetup{
 pdfauthor={Johannes Passing},
 pdftitle={Dynamic Tracing of Windows NT Kernel Mode Components},
 pdfdisplaydoctitle=true,
 pdfsubject={NTrace},
 pdfkeywords={Windows, NT, Tracing, Debugging, WRK},
 bookmarksnumbered=false
 }
 
%% Define a new 'leo' style for the package that will use a smaller font.
\makeatletter
\def\url@leostyle{%
  \@ifundefined{selectfont}{\def\UrlFont{\sf}}{\def\UrlFont{\small\ttfamily}}}
\makeatother
%% Now actually use the newly defined style.
\urlstyle{leostyle}


\sloppy
\setcounter{tocdepth}{4}

\renewcommand\floatpagefraction{.9}
\renewcommand\topfraction{.9}
\renewcommand\bottomfraction{.9}
\renewcommand\textfraction{.1}   
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

\begin{document}

%% ++++++++++++++++++++++++++++++++++++++++++
%% page number tweaking
%% + openany option
%% - font size 11 (?)
%% ++++++++++++++++++++++++++++++++++++++++++


%% ++++++++++++++++++++++++++++++++++++++++++
%% frontmatter
%% ++++++++++++++++++++++++++++++++++++++++++
%\frontmatter
\pagenumbering{roman} 

\include{thesis_titlepage}
\include{thesis_abstract}
\include{thesis_acknowledgments}

%% ++++++++++++++++++++++++++++++++++++++++++
%% Verzeichnisse
%% ++++++++++++++++++++++++++++++++++++++++++
\ifnotdraft{
\tableofcontents
\listoffigures
\listoftables
\lstlistoflistings
}



%% ++++++++++++++++++++++++++++++++++++++++++
%% Hauptteil
%% ++++++++++++++++++++++++++++++++++++++++++

\mainmatter
\pagenumbering{arabic}

%----------------------------------------------------------------------
%----------------------------------------------------------------------
\part{Theoretical Groundwork}
\label{part:TheoreticalGroundwork}
\chapter{Introduction}

With growing complexity, software systems have become increasingly hard to understand 
and debug. Confronted with large volumes of source code, it is challenging for a developer to
reason about the potential behavior of a system at runtime, much less to diagnose erroneous 
behavior exhibited by a running system. Having appropriate tools at hand is 
hence indispensable.

This thesis covers dynamic tracing, a building block for creating tools helping to
analyze systems by observing their bevavior. Observing a program as it is being executed, 
and not requiring it to be rebuilt or restarted, dynamic tracing
is particularly well-suited for long running systems such as operating system kernels. 
The focus of this thesis therefore lies on the analysis of the Windows NT kernel and its components.

The topic can be further refined as addressing function boundary tracing, i.e. capturing
function entry and exit events. Software tracing can be performed at an even more fine-grained level,
yet, function boundary tracing may be expected to yield rather universally applicable results for
the aforementioned purposes.



\section{Potential Fields of Applications}
Given this background, potential and appropriate fields of applications for a dynamic 
tracing system for the Windows NT kernel can now be laid out more clearly. 

An important potential field of application is production debugging. In production debugging scenarios,
a dynamic tracing solution promises to allow instrumentation and recording of information without
interrupting the service of the examined system. Once information has been collected,
instrumentation can be revoked and the data can be analyzed \emph{offline}. 

Besides the capability of capturing information valuable to pinpoint flaws in a software 
system, a dynamic function boundary tracing system also promises to help in comprehending a 
system. Traces, in particular when visualized properly -- for instance, as call graphs --
can help illuminating the runtime behavior and dynamics of a software system. 

Dynamic tracing can also be used for profiling. In contrast to sampling-based profilers 
such as \emph{kernrate}\footnote{Part of the Microsoft Windows Server 2003 Resource Kit}, 
such a profiler could potentially be more accurate as the risk of \emph{missing} a 
function call due to the sampling interval chosen ceases to exist.

In a similar manner, code coverage -- albeit limited to function level rather than
basic block level -- could be implemented based on such a dynamic tracing solution.

%\section{Problem Statement}
%Having sketched out the role of debugging, static and dynamic tracing, it is
%now possible to refine the problems addressed by this work.
%
%The overall aim of this thesis is to assess the feasibility and viablility 
%of dynamic tracing for kernel mode Windows. This includes
%
%\begin{itemize}
%	\item Analyzing the state of the art of dynamic tracing: A significant
%				amount of reseach has been performed on the topic of dynamic tracing.
%				Moreover, a wide range of dynamic tracing approaches have already been developed
%				in the past. Analyzing strengths and weaknesses of existing approaches and 
%				discussing their applicability to kernel mode Windows is thus crucial.
%	\item Defining a set of criteria a prospective tracing solution is to meet.
%	\item Developing a dynamic tracing approach tailored to kernel mode Windows.
%	\item Proving feasibility and viablility of this approach by an implementation.
%	\item Analyzing the implementation for meeting the criteria set up.
%\end{itemize}

\section{Structure of this Thesis}
The thesis is structured into three parts. Part 1 covers the theoretical groundwork of the topic.
This includes an assessment and classification of existing dynamic tracing approaches, along with 
a discussion of their strengths and weaknesses. The first part concludes with a detailed discussion
of common implementation challenges related to runtime code modification-based tracing approaches.

Part II introduces NTrace, a dynamic tracing solution developed as part of this thesis. The part opens
with a general discussion of the tool architecture and continues by laying out the basic approach to function boundary tracing 
taken. Based on this groundwork, the details of the approach as well as the implementation challenges 
are discussed.

The thesis concludes with Part III, which provides an analysis of NTrace. This includes a discussion and measurements 
of the performance characteristics of the implementation.


\section{Contributions}

The main contributions of this work are as follows: 
\begin{itemize}
	\item Definition of a classification scheme for dynamic tracing techniques.
	\item Proposal of a dynamic function boundary tracing system for the Windows NT kernel that fully integrates with
				the Windows Structured Exception Handling infrastructure.
				
%	\item Demonstration of that the Microsoft Windows Hotpatching infrastructure can be 
%			  leveraged for dynamic tracing purposes.
%	\item Advancement the technique of using return address replacement and 
%			  auxiliary stack usage for tracing so that it can now effectively be deployed in kernel mode
%			  Windows.
%	\item Development of a mechanism for instrumenting the i386 \emph{Structured Exception Handling}
%				infrastructure.
	\item Proving the feasibility of the proposed system by an implementation and evaluating its
				runtime performance.
\end{itemize}


\section{Definition of Terms}
\label{sec:DefinitionOfTerms}
The semantic differences between the terms function, procedure, and routine are insignificant 
in the context of this work and the terms are used interchangeably. Following
\cite{unix03}, the terms \emph{module} (or \emph{executable module}) and \emph{image} (or \emph{program image})
are distinguished as follows: A module constitutes a file, usually generated by a linker, containing executable
code. When the module is loaded into a process, the in-memory representation of the module 
is referred to as \emph{image}.

For the remainder of this work, code is considered \emph{unmodified} if it matches the code
that would be executed in the absence of a tracing system.

The terms \emph{trampoline}, \emph{springboard} and \emph{bounce}, all commonly encountered
in the literature 
%\cite{Brubacher99} \cite{hollingsworth94dynamic} \cite{Bruening04} \cite{Pearce02} \cite{tamches99finegrained}, 
\citep{Brubacher99,hollingsworth94dynamic,Bruening04,Pearce02,tamches99finegrained}, 
will be jointly referred to as \emph{trampoline}.

Any machine that is capable of concurrently executing more than one thread 
is referred to as multiprocessor machine. This includes true multiprocessor machines as well
as machines using multicore CPUs or CPUs supporting technologies such as 
\emph{Simultaneous Multithreading} \cite{Eggers_1simultaneous}.

The terms \emph{i386} or \emph{x86} are avoided in favor of the term \emph{IA-32} 
\cite{intel07_1} to denote the Intel 32 bit architecture. The \emph{AMD64} \cite{amd64_2}
and \emph{Intel64} \cite{intel07_1} architectures are jointly referred to as \emph{AMD64}.

All references to Windows refer to the Windows NT family exclusively. Unless stated 
otherwise, all discussions of Windows NT further refer to Windows Server 2003 SP1/SP2 
and the Windows Research Kernel 1.2 only. The term \emph{kernel} as used in this work 
does not include drivers but only refers to the functionality implemented in the 
kernel module, usually named ntoskrnl.exe. In contrast, \emph{kernel mode Windows}
refers to the entirety of code running in kernel mode.

Unless stated otherwise, all references to \emph{compiler} and \emph{linker} shall 
denote the Microsoft compiler \emph{cl} and linker \emph{link}, respectively. 
As Microsoft compilers and linkers are the prevalent tools for kernel mode Windows 
development, other compilers are not addressed by this work. 

Occasionally, use of the syntax \verb|module!Function| is made to fully qualify 
routine names. Finally, all assembler code listings use the Intel syntax. 

\chapter{Criteria}
\label{sec:Criteria}
To allow a more effective assessment of existing and potential tracing techniques,
a set of basic criteria such techniques are to meet shall be defined.

In their paper \emph{Dynamic Program Instrumentation for Scalable Performance Tools},
Hollingsworth et al. \cite{hollingsworth94dynamic} state that monitoring the performance of 
massively parallel programs requires an instrumentation system that is \emph{detailed}, 
\emph{frugal}, and \emph{scalable}.

The focus of this work is not massively parallel systems but rather general 
purpose systems -- still, these three criteria remain valid and provide a suitable
foundation. However, to emphasize the fact that monitoring or tracing activity should not endanger
system stability, \emph{Robustness} shall be introduced as a fourth criterion.

\section*{Detailed}
In the realm of function boundary tracing, the desired level of detail is
defined by function calls. A solution capable of tracing function entries
can be expected to already produce valuable results. Yet, it is desirable to
have the tracing solution also support function exit tracing, as this additionally allows
inferring call relationships among functions. 

Whether more fine grained information such as parameter and return values are of
importance depends on the individual purpose. If the intent of tracing is to
collect performance information or to comprehend the overall system, such 
additional information may be not of interest. For problem analysis, in contrast,
this additional information may be valuable indeed. 

To keep the discussion general in this regard, the aspects of more fine grained 
information are therefore ignored in the remainder of this work.

\section*{Frugal}
Frugality -- and performance in particular -- is at the core of a dynamic tracing solution. 
In fact, the promise of low runtime overhead may have been the reason for 
a user deciding for a dynamic tracing solution in the first place. More generally, economical
usage of resources is important for limiting the impact on resource and timing behavior of 
the traced system, and thus, to ensure applicability of the tracing system.

Rather than striving at high efficiency, frugality can also be attained by limiting the amount
of data collected. This, however, may stand in conflict with the demand to be detailed. As such,
this conflict depicts the classical conundrum of having to balance between performance and precision.

Another aspect of frugality is the runtime overhead in case all tracing activity has been 
disabled. That is, the tracing system is active, yet no tracing information is currently being
captured. In this case, it is desirable for the tracing solution to impose no or at least 
no significant runtime overhead.

\section*{Scalable}
Scalability describes the ability of the tracing system to remain frugal
and detailed when resources are added. The most important resources in this regard are additional
functions to be traced, and additional processors.

When tracing information is to be captured for a growing number of functions, the amount
of data that is to be handled will grow as well. To remain frugal, the system must be able
to efficiently handle these increased amounts of data.

As multiprocessor machines -- and multicore processors in particular -- become increasingly
common, the ability to properly deal with more than one processor becomes crucial as well.
Certainly, this requires the tracing system to be written in a multiprocessor-safe manner.
However, to remain frugal, this also requires that the tracing system makes effective use
of the additional processors. In particular, it should not limit the degree in which the 
tracee makes use of the additional processors and it should not impose significant performance 
bottlenecks itself.

\section*{Robust}
When tracing is used in development scenarios, robustness may not be of utmost concern. In
production debugging scenarios, however, it is crucial that tracing does not endanger
stability and availability of the traced system. 

Runtime code modification, a technique commonly used for implementing dynamic tracing, 
is particularly sensitive in this regard. As will be covered in more depth in the remainder
of this work, significant attention to the details have to be paid in order not to undermine
the stability of the system. 

%Especially in the realm of Windows, there have been countless attempts of using runtime code 
%modification -- both by independent developers and commercially \cite{Skywing06} -- that have 
%failed to properly address these issues. 
%It therefore does not come at a surprise that in  
%newsgroups such as \emph{ntdev}, run by Open System Resources, a general aversion towards 
%such practice can be observed. 

As such, robustness clearly is an important factor to consider when assessing dynamic tracing
solutions. 


%\section*{Windows NT Kernel Specifics}
%With focus on Windows NT, a number of additional, more specific crtiteria can
%be identified. 
%
%\subsection*{Debugging Symbols}
%Acting on binary code, dynamic instrumentation solutions usually need appropriate
%symbolic information in order to allow proper instrumentation. If source code is available
%for the software to be traced, generating appropriate metadata is feasible. In the case
%of the Windows kernel and drivers, however, source code must be assumed to be not available.
%Rather, usage of symbolic information must be restricted to the information Microsoft provides.
%
%Indeed, to allow more efficient debugging, Microsoft provides debugging symbol information 
%for the vast majority of binaries, including kernel and drivers. The point to notice, however,
%is that this information comprises \emph{public symbols} only -- that is, only a limited set
%of symbolic information is provided. For a tracing system that aims to be effective for the 
%Windows kernel, this means that it must be able to cope with the limited amount of 
%symbolic information available.
%
%\subsection*{Exception Handling}
%With \emph{Structured Exception Handling}, the Windows NT kernel has builtin support 
%for software exceptions. Although not strictly related to function boundary tracing per se,
%exceptions can have significant impact on the control flow of a program. As such, it
%is desireable to have a dynamic tracing solution recognize exceptions and treat them
%appropriately.
%

%Motivate Multiproc?
%
% - no disasm
% - uninstrumentability
% - dynamically loadable, unloadable
% 


%\chapter{Dynamic Tracing}
%
%...
%
%
%Performing these steps at runtime and relying on unmodified binaries frees the 
%user from having to decide between different binaries before launching the
%program -- a decision that has to be made in the case of static tracing approaches. 
%Moreover, dynamic tracing solutions usually enable the user to repeatedly alter
%his decisions about what is to be traced in which level of detail. Being able
%to refine or modify the tracing activity at runtime in turn promises to allow
%the user to obtain the insights sought after more quickly than in the case of static tracing.
%
%Furthermore, as long as no tracing activity is taking place, most dynamic tracing 
%solutions can be expected not to induce any runtime overhead. 
%
%% disadvantages: less flexible in what is to be traced?


%\section{Related Approaches}

%\section{Windows NT Builtins}
%% ETW
%\section{Static Instrumentation}
%\subsection{Manual Source Code Augmentation}
% LTT, ETW
% -> olszewski, 2.1
%\subsection{Compiler Level Augmentation}
%\subsection{Binary Rewriting}

% ATOM, Purify, EEL, PLTO, Pixie, ETCH
% -> olszewski, 2.1

% SPIN, Vino, Exokernel
%-> tamches diss §2.1.1


\include{thesis_classification}
\include{thesis_classanalysis}
%\include{thesis_existing}
\include{thesis_codemod}
\include{thesis_ownapproach}



%% ++++++++++++++++++++++++++++++++++++++++++
%% Anhang
%% ++++++++++++++++++++++++++++++++++++++++++

\appendix



%% ++++++++++++++++++++++++++++++++++++++++++
%% Literatur
%% ++++++++++++++++++++++++++++++++++++++++++
%  mit dem Befehl \nocite werden auch nicht 
%  zitierte Referenzen abgedruckt
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{\bibname}
%\nocite{*} 
\bibliographystyle{alpha}
\bibliography{bibfile}

%% ++++++++++++++++++++++++++++++++++++++++++
%% Index
%% ++++++++++++++++++++++++++++++++++++++++++
\ifnotdraft{
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Index}
\printindex            % Index, Stichwortverzeichnis
}
\include{thesis_confm}
\end{document}
%% end of file

